---
title: "STAT5205 Final Project"
author: "Tingjun Kang, Yuhao Wang, Linke Wu"
date: "15 December 2023"
output:
  pdf_document: default
  latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(tidyverse) 
library(car)
library(MASS)
wholecoal <- read_csv("coal_hhv.csv")
# Load the tidyverse package so it is available to use
# Load the Car package so it is available to use
# Import data from document
```


## EDA for total data

```{r}
newcoal <- wholecoal %>%
  filter(!is.na(Ash_wt) & !is.na(Moisture_wt) & !is.na(Volatile_Matter_wt) & !is.na(Fixed_Carbon_wt) & !is.na(HHV))
# Data cleaning
# Remove missing values on each value
```

```{r}
glimpse(newcoal)
summary(newcoal)
# Numerical summaries of each variable
```
```{r}
IQR(newcoal$Moisture_wt)
IQR(newcoal$Ash_wt)
IQR(newcoal$Volatile_Matter_wt)
IQR(newcoal$Fixed_Carbon_wt)
IQR(newcoal$HHV)
# Calculate IQR of each variable
```
```{r}
sd(newcoal$Moisture_wt)
sd(newcoal$Ash_wt)
sd(newcoal$Volatile_Matter_wt)
sd(newcoal$Fixed_Carbon_wt)
sd(newcoal$HHV)
# Calculate standard deviation of each variable
```
```{r}
ggplot(data = newcoal, aes(x=HHV)) +
  geom_histogram(fill="coral2",color="grey",bins = 20) +
  labs(x="Higher Heating Value", title = "Histogram of HHV")
# Create a histogram for Higher Heating Value
```

```{r}
ggplot(data=newcoal, aes(x=Moisture_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Moisture", y="Higher Heating Value",title = "Higher Heating Value with Moisture")
```

```{r}
ggplot(data=newcoal, aes(x=Ash_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Ash", y="Higher Heating Value",title = "Higher Heating Value with Ash")
```
```{r}
ggplot(data=newcoal, aes(x=Volatile_Matter_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Volatile Matter", y="Higher Heating Value",title = "Higher Heating Value with Volatile Matter")
```
```{r}
ggplot(data=newcoal, aes(x=Fixed_Carbon_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Fixed Carbon", y="Higher Heating Value",title = "Higher Heating Value with Fixed Carbon")
```

##Divide data into training set and testing set:

```{r}
set.seed(923)
# Create the traning and test set
```

```{r}
n <- nrow(wholecoal)
# record the number of observations in dataset
```

```{r}
train_ind <- sample(1:n, size = round(0.8*n))
# choosing 80% observations as training in random
```

```{r}
train <- wholecoal %>% filter(No.coal %in% train_ind)
# Create a training set
```

```{r}
test <- wholecoal %>% filter(!(No.coal %in% train_ind))
# Create a testing set
```

##Processing EDA of training set:

```{r}
newcoal <- train %>%
  filter(!is.na(Ash_wt) & !is.na(Moisture_wt) & !is.na(Volatile_Matter_wt) & !is.na(Fixed_Carbon_wt) & !is.na(HHV))
# Data cleaning
# Remove missing values on each value
```

```{r}
glimpse(newcoal)
summary(newcoal)
# Numerical summaries of each variable
```

```{r}
IQR(newcoal$Moisture_wt)
IQR(newcoal$Ash_wt)
IQR(newcoal$Volatile_Matter_wt)
IQR(newcoal$Fixed_Carbon_wt)
IQR(newcoal$HHV)
# Calculate IQR of each variable
```

```{r}
sd(newcoal$Moisture_wt)
sd(newcoal$Ash_wt)
sd(newcoal$Volatile_Matter_wt)
sd(newcoal$Fixed_Carbon_wt)
sd(newcoal$HHV)
# Calculate standard deviation of each variable
```

```{r}
ggplot(data = newcoal, aes(x=HHV)) +
  geom_histogram(fill="coral2",color="grey",bins = 20) +
  labs(x="Higher Heating Value", title = "Histogram of HHV")
# Create a histogram for Higher Heating Value
```

```{r}
ggplot(data=newcoal, aes(x=Moisture_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Moisture", y="Higher Heating Value",title = "Higher Heating Value with Moisture")
# Create a scatter plots between "HHV" and "Moisture"

```

```{r}
ggplot(data=newcoal, aes(x=Ash_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Ash", y="Higher Heating Value",title = "Higher Heating Value with Ash")
# Create a scatter plots between "HHV" and "Ash_wt"
```

```{r}
ggplot(data=newcoal, aes(x=Volatile_Matter_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Volatile Matter", y="Higher Heating Value",title = "Higher Heating Value with Volatile Matter")
# Create a scatter plots between "HHV" and "Volatile_Matter_wt"
```

```{r}
ggplot(data=newcoal, aes(x=Fixed_Carbon_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Fixed Carbon", y="Higher Heating Value",title = "Higher Heating Value with Fixed Carbon")
# Create a scatter plots between "HHV" and "Fixed_Carbon_wt"
```


##Model buliding:

```{r}
full_model <- lm(HHV ~ Moisture_wt+Fixed_Carbon_wt+Volatile_Matter_wt+Ash_wt,data=train)
vif(full_model)
# Bulid the full_model with all predictors 
# Check VIF then decide which predictor needs removed 
```


```{r}
model_1 <- lm(HHV~Ash_wt+Fixed_Carbon_wt+Volatile_Matter_wt,data=train)
vif(model_1)
# Remove the "Moisture_wt" since it has the largest VIF
# Bulid model_1 then calcuate VIF
```

```{r}
summary(model_1)
# Select model_1 as the manual selected model
```


```{r}
stepAIC(full_model,direction = "both",k=2)
# Use automated stepwise selection to reduce full model
```

```{r}
model_2 <- lm(HHV ~ Ash_wt + Fixed_Carbon_wt + Moisture_wt,data=train)
summary(model_2)
vif(model_2)
# Bulid model_2 based on automated stepwise selected model
```



```{r}
select_criteria = function(model, n)
{
  SSres <- sum(model$residuals^2)
  Rsq_adj <- summary(model)$adj.r.squared
  p <- length(model$coefficients) - 1
  AIC <- n*log(SSres/n) + 2*p
  AICc <- AIC + (2*(p+2)*(p+3)/(n-p-1))
  BIC <- n*log(SSres/n) + (p+2)*log(n)
  res <- c(SSres, Rsq_adj, AIC, AICc, BIC)
  names(res) <- c("SSres", "Rsq_adj", "AIC", "AIC_c", "BIC")
  return(res)
}
select_criteria(model_1,40)
select_criteria(model_2,40)
#compare manual selected model and automated stepwise selected model
```

##Model diagnostics:
#For model_1

```{r}
pairs(~ Fixed_Carbon_wt + Volatile_Matter_wt + Ash_wt,data=train,main="Pair Plots of Predictors for Model_1")
# Check multicollinearity with pair plots of predictors for model_1
```


#For model_2

```{r}
pairs(~ Moisture_wt + Fixed_Carbon_wt + Ash_wt,data=train,main="Pair Plots of Predictors for Model_2")
# Check multicollinearity with pair plots of predictors for model_2
```

##Check residual plot and residual Q-Q plot:
#For model_1
```{r}
Res <- rstandard(model_1)
Fit <- fitted(model_1)
plot(Res~Fit,main="Residual vs fitted value for model_1")
# Residual vs fitted value for model_1
```

```{r}
par(mfrow = c(1,3))
plot(Res~train$Ash_wt,xlab="Ash_wt",main="Residual vs Ash_wt")
plot(Res~train$Fixed_Carbon_wt, xlab="Fixed_Carbon_wt",main="Residual vs Fixed_Carbon_wt")
plot(Res~train$Volatile_Matter_wt, xlab="Volatile_Matter_wt",main="Residual vs Volatile_Matter_wt")
# Residual vs Predictors for model_1
```

```{r}
qqnorm(Res)
qqline(Res)
# Residual Q-Q plot for model_1
```

#For model_2
```{r}
Residuals <- rstandard(model_2)
Fitted <- fitted(model_2)
plot(Residuals~Fitted,main="Residual vs fitted value for model_2")
# Residual vs fitted value for model_2
```

```{r}
par(mfrow = c(1,3))
plot(Residuals~train$Ash_wt,xlab="Ash_wt",main="Residual vs Ash_wt")
plot(Residuals~train$Fixed_Carbon_wt, xlab="Fixed_Carbon_wt",main="Residual vs Fixed_Carbon_wt")
plot(Residuals~train$Moisture_wt, xlab="Moisture_wt",main="Residual vs Moisture_wt")
# Residual vs Predictors for model_2
```

```{r}
qqnorm(Residuals)
qqline(Residuals)
# Residual Q-Q plot for model_2
```

##Determine the Leverage points, outliers,influential points:
#For model_1
```{r}
h <- hatvalues(model_1)
threshold <- 2*(length(model_1$coefficients)/nrow(train))
L <- which(h > threshold)
train[L,]
# Determine whether there are Leverage points
```

```{r}
std_res <- rstandard(model_1)
O <- which(abs(std_res)>2)
train[O,]
# Determine whether there are outliers
```

```{r}
D <- cooks.distance(model_1)
cutoff <- qf(0.5, length(model_1$coefficients), nrow(train)-length(model_1$coefficients), lower.tail=T)
I <- which(D > cutoff)
train[I,]
# Determine whether there are influential points
```

#For model_2
```{r}
h <- hatvalues(model_2)
threshold <- 2*(length(model_2$coefficients)/nrow(train))
L <- which(h > threshold)
train[L,]
# Determine whether there are Leverage points
```

```{r}
std_res <- rstandard(model_2)
O <- which(abs(std_res)>2)
train[O,]
# Determine whether there are outliers
```

```{r}
D <- cooks.distance(model_2)
cutoff <- qf(0.5, length(model_2$coefficients), nrow(train)-length(model_2$coefficients), lower.tail=T)
I <- which(D > cutoff)
train[I,]
# Determine whether there are influential points
```

##Model Validation2:

#Processing EDA of testing set:(compare EDA)

```{r}
newcoal <- test %>%
  filter(!is.na(Ash_wt) & !is.na(Moisture_wt) & !is.na(Volatile_Matter_wt) & !is.na(Fixed_Carbon_wt) & !is.na(HHV))
# Data cleaning
# Remove missing values on each value
```

```{r}
glimpse(newcoal)
summary(newcoal)
# Numerical summaries of each variable
```

```{r}
IQR(newcoal$Moisture_wt)
IQR(newcoal$Ash_wt)
IQR(newcoal$Volatile_Matter_wt)
IQR(newcoal$Fixed_Carbon_wt)
IQR(newcoal$HHV)
# Calculate IQR of each variable
```

```{r}
sd(newcoal$Moisture_wt)
sd(newcoal$Ash_wt)
sd(newcoal$Volatile_Matter_wt)
sd(newcoal$Fixed_Carbon_wt)
sd(newcoal$HHV)
# Calculate standard deviation of each variable
```

```{r}
ggplot(data = newcoal, aes(x=HHV)) +
  geom_histogram(fill="coral2",color="grey",bins = 20) +
  labs(x="Higher Heating Value", title = "Histogram of HHV")
# Create a histogram for Higher Heating Value
```

```{r}
ggplot(data=newcoal, aes(x=Moisture_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Moisture", y="Higher Heating Value",title = "Higher Heating Value with Moisture")
# Create a scatter plots between "HHV" and "Moisture"

```

```{r}
ggplot(data=newcoal, aes(x=Ash_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Ash", y="Higher Heating Value",title = "Higher Heating Value with Ash")
# Create a scatter plots between "HHV" and "Ash_wt"
```

```{r}
ggplot(data=newcoal, aes(x=Volatile_Matter_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Volatile Matter", y="Higher Heating Value",title = "Higher Heating Value with Volatile Matter")
# Create a scatter plots between "HHV" and "Volatile_Matter_wt"
```

```{r}
ggplot(data=newcoal, aes(x=Fixed_Carbon_wt, y=HHV)) +
  geom_point() +
  geom_smooth(method = lm,se=FALSE) +
  labs(x="Fixed Carbon", y="Higher Heating Value",title = "Higher Heating Value with Fixed Carbon")
# Create a scatter plots between "HHV" and "Fixed_Carbon_wt"
```

#Refit 2 models on testing data

```{r}
model_1_test <- lm(HHV~Ash_wt+Fixed_Carbon_wt+Volatile_Matter_wt,data=test)
summary(model_1_test)
# Refit model_1 with testing data
```

```{r}
model_2_test <- lm(HHV ~ Ash_wt + Fixed_Carbon_wt + Moisture_wt,data=test)
summary(model_2_test)
# Refit model_2 with testing data
```

#check 2 new models residual plot and residual Q-Q plot:
#For model_1_test
```{r}
Res <- rstandard(model_1_test)
Fit <- fitted(model_1_test)
plot(Res~Fit,main="Residual vs fitted value for model_1_test")
# Residual vs fitted value for model_1_test
```

```{r}
par(mfrow = c(1,3))
plot(Res~test$Ash_wt,xlab="Ash_wt",main="Residual vs Ash_wt")
plot(Res~test$Fixed_Carbon_wt, xlab="Fixed_Carbon_wt",main="Residual vs Fixed_Carbon_wt")
plot(Res~test$Volatile_Matter_wt, xlab="Volatile_Matter_wt",main="Residual vs Volatile_Matter_wt")
# Residual vs Predictors for model_1_test
```

```{r}
qqnorm(Res)
qqline(Res)
# Residual Q-Q plot for model_1_test
```

#For model_2_test
```{r}
Residuals <- rstandard(model_2_test)
Fitted <- fitted(model_2_test)
plot(Residuals~Fitted,main="Residual vs fitted value for model_2_test")
# Residual vs fitted value for model_2_test
```

```{r}
par(mfrow = c(1,3))
plot(Residuals~test$Ash_wt,xlab="Ash_wt",main="Residual vs Ash_wt")
plot(Residuals~test$Fixed_Carbon_wt, xlab="Fixed_Carbon_wt",main="Residual vs Fixed_Carbon_wt")
plot(Residuals~test$Moisture_wt, xlab="Moisture_wt",main="Residual vs Moisture_wt")
# Residual vs Predictors for model_2_test
```

```{r}
qqnorm(Residuals)
qqline(Residuals)
# Residual Q-Q plot for model_2_test
```

##Determine the Leverage points, outliers,influential points of 2 new model:
#For model_1_test
```{r}
h <- hatvalues(model_1_test)
threshold <- 2*(length(model_1_test$coefficients)/nrow(test))
L <- which(h > threshold)
train[L,]
# Determine whether there are Leverage points
```

```{r}
std_res <- rstandard(model_1_test)
O <- which(abs(std_res)>2)
train[O,]
# Determine whether there are outliers
```

```{r}
D <- cooks.distance(model_1_test)
cutoff <- qf(0.5, length(model_1_test$coefficients), nrow(test)-length(model_1_test$coefficients), lower.tail=T)
I <- which(D > cutoff)
train[I,]
# Determine whether there are influential points
```

#For model_2_test
```{r}
h <- hatvalues(model_2_test)
threshold <- 2*(length(model_2_test$coefficients)/nrow(test))
L <- which(h > threshold)
train[L,]
# Determine whether there are Leverage points
```

```{r}
std_res <- rstandard(model_2_test)
O <- which(abs(std_res)>2)
train[O,]
# Determine whether there are outliers
```

```{r}
D <- cooks.distance(model_2_test)
cutoff <- qf(0.5, length(model_2_test$coefficients), nrow(test)-length(model_2_test$coefficients), lower.tail=T)
I <- which(D > cutoff)
train[I,]
# Determine whether there are influential points
```

##calculate Adj.R^2 and AIC,BIC for testing model
```{r}
select_criteria = function(model, n)
{
  SSres <- sum(model$residuals^2)
  Rsq_adj <- summary(model)$adj.r.squared
  p <- length(model$coefficients) - 1
  AIC <- n*log(SSres/n) + 2*p
  AICc <- AIC + (2*(p+2)*(p+3)/(n-p-1))
  BIC <- n*log(SSres/n) + (p+2)*log(n)
  res <- c(SSres, Rsq_adj, AIC, AICc, BIC)
  names(res) <- c("SSres", "Rsq_adj", "AIC", "AIC_c", "BIC")
  return(res)
}
select_criteria(model_1_test,7)
select_criteria(model_2_test,7)
#calculate ADJ.R^2 and AIC,BIC for testing model
```



#Final model and Data prediction:

```{r}
# Choose model_2 and do the prediction
prediction <- predict(model_2,test)
prediction
sum((prediction-test$HHV)^2)
# Calculate sse of prediction
```


```{r}
# Choose model_1 and do the prediction
prediction <- predict(model_1,test)
prediction 
sum((prediction-test$HHV)^2)
# Calculate sse of prediction
```
 
##Choose model_1 as the final model

